Appendix B: Exploratory LPA for Ocean Microplastics

### 2023-01-23 appendix_b.tex
edits by Mauricio

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Appendix A: Confirmatory LPA for the Caregiver Compass},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Appendix A: Confirmatory LPA for the Caregiver Compass}
\author{\phantom{0}}
\date{}

@@

This is an example of confirmatory LPA using \texttt{tidySEM}.
The simulated data are based on work by Zegwaard and colleagues, who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
Qualitative research among experts resulted in a theory postulating the existence of four types of caregivers (translated from the original Dutch):

\textbf{Balanced}

\begin{quote}
The balanced caregiver experiences relative balance between the costs and benefits of caring for a close other.
\end{quote}

\textbf{Imbalanced}

\begin{quote}
The imbalanced caregiver experiences a precarious balance between the costs and benefits of caring for a close other.
\end{quote}

\textbf{Lonely}

\begin{quote}
The lonely caregiver experiences a strong sense of isolation.
\end{quote}

\textbf{Entrapped}

\begin{quote}
The entrapped caregiver strongly feels a sense of being entangled in responsibilities which are difficult to fulfill.
\end{quote}

To view the data documentation, run the command \texttt{?tidySEM::zegwaard\_carecompass} in the R console.

@@

To load the data, simply attach the \texttt{tidySEM} package.
For convenience, we assign the variables used for analysis to an object called \texttt{df}.
We first only use the four scales: \texttt{c("burdened",\ "trapped",\ "negaffect",\ "loneliness")}.
`.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Load required packages}
\FunctionTok{library}\NormalTok{(tidySEM) }
\FunctionTok{library}\NormalTok{(ggplot2)}
\CommentTok{\# Load data}
\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ zegwaard\_carecompass[, }\FunctionTok{c}\NormalTok{(}\StringTok{"burdened"}\NormalTok{, }\StringTok{"trapped"}\NormalTok{, }\StringTok{"negaffect"}\NormalTok{, }\StringTok{"loneliness"}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

@@

As per the best practices, the first step in LCA is examining the observed data.
We use \texttt{tidySEM::descriptives()} to describe the data numerically.
Because all scales are continuous, we select only columns for continuous data to de-clutter the table:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ tidySEM}\SpecialCharTok{::}\FunctionTok{descriptives}\NormalTok{(df)}
\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{,}
                 \StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{,}
                 \StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
\NormalTok{desc}
\end{Highlighting}
\end{Shaded}

@@

\begin{tabular}{lllllllllll}
\toprule
name & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{missing} & \multicolumn{1}{c}{unique} & \multicolumn{1}{c}{mean} & \multicolumn{1}{c}{median} & \multicolumn{1}{c}{sd} & \multicolumn{1}{c}{min} & \multicolumn{1}{c}{max} & \multicolumn{1}{c}{skew\_2se} & \multicolumn{1}{c}{kurt\_2se}\\
\midrule
burdened & 509 & 0.01 & 509 & 3.38 & 3.39 & 0.75 & 1.20 & 5.28 & 0.17 & 6.51\\
trapped & 505 & 0.02 & 505 & 1.73 & 1.79 & 0.90 & -0.86 & 3.81 & -1.03 & 5.39\\
negaffect & 506 & 0.01 & 506 & 2.50 & 2.52 & 0.69 & 0.71 & 4.98 & 0.08 & 6.52\\
loneliness & 510 & 0.01 & 510 & 2.66 & 2.69 & 0.62 & 0.98 & 4.22 & -0.33 & 6.31\\
\bottomrule
\end{tabular}

@@

The table indicates two potential causes for concern: there is a small percentage of missingness, and all variables have relatively high kurtosis.
Since there are some missing values, we can conduct an MCAR test using \texttt{mice::mcar(df)}.
According to Hawkins' test, there is no evidence to reject the assumptions of multivariate normality and MCAR, \(\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71\).
Missing data will be accounted for using FIML.

Additionally, we can plot the data.
The \texttt{ggplot2} function \texttt{geom\_density()} is useful for continuous data.
Visual inspection confirms the conclusions from the \texttt{descriptives()} table: the data are kurtotic (peaked).

@@

\includegraphics{appendices/plot_lpa_desc.pdf}

@@

As all variables are continuous, we can use the convenience function \texttt{tidySEM::mx\_profiles()}, which is a wrapper for the generic function \texttt{mx\_mixture()} optimized for continuous indicators.
Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
Its arguments are \texttt{data} and number of \texttt{classes}.
All variables in \texttt{data} are included in the analysis, which is why we first selected the indicator variables.
As this is a confirmatory LCA, we do not follow a strictly data-driven class enumeration procedure.
We will set the maximum number of classes \(K\) to one more than the theoretically expected number.
We set a seed to ensure replicable results.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df,}
                   \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

This analysis should produce some warnings about cluster initialization.
These relate to the selection of starting values, which relies on the K-means algorithm and is not robust to missing data.
The algorithm automatically switches to hierarchical clustering, no further action is required.

@@

To compare the fit of the theoretical model against other models, we create a model fit table using \texttt{table\_fit()} and retain relevant columns.
We also determine whether any models can be disqualified.

In this example, all models converge without issues.
If, for example, the two-class solution had not converged, we could use the function \texttt{res{[}{[}2{]}{]}\ \textless{}-\ mxTryHard(res{[}{[}2{]}{]})} to aid convergence.

Next, we check for local identifiability.
The sample size is consistently reported as 513, which means that partially missing cases were indeed included via FIML.
The smallest class size occurs in the 5-class model, where the smallest class is assigned 7\% of cases, or 38 cases.
This model has 28 parameters, approximately 6 per class.
We thus have at least five observations per parameter in every class, and do not disqualify the 5-class model.

There are concerns about theoretical interpretability of all solutions, as the entropies and minimum classification probabilities are all low.
However, in this confirmatory use case, we address this when interpreting the results.

@@

\caption{\label{tab:tabfit}Model fit table}

\begin{tabular}{lllllllllll}
\toprule
Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{p} & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Entropy} & \multicolumn{1}{c}{p\_min} & \multicolumn{1}{c}{p\_max} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{n\_max} & \multicolumn{1}{c}{lmr\_p}\\
\midrule
equal var 1 & -2,241.98 & 8 & 513 & 4,533.89 & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 & NA\\
equal var 2 & -2,031.37 & 13 & 513 & 4,143.85 & 0.74 & 0.91 & 0.93 & 0.42 & 0.58 & 0.00\\
equal var 3 & -1,951.35 & 18 & 513 & 4,015.02 & 0.78 & 0.89 & 0.91 & 0.19 & 0.54 & 0.00\\
equal var 4 & -1,916.25 & 23 & 513 & 3,976.03 & 0.75 & 0.81 & 0.92 & 0.16 & 0.34 & 0.00\\
equal var 5 & -1,912.93 & 28 & 513 & 4,000.59 & 0.70 & 0.39 & 0.92 & 0.07 & 0.31 & 0.28\\
\bottomrule
\end{tabular}

@@

Class enumeration unambiguously indicates a preference for the 4-class solution: the 4-class solution has the lowest BIC, which means it is preferred over all other solutions including a 1-class solution and a solution with more classes.
The p-values of the LMR LR test are significant for all pairwise model comparisons, except for the 5-class model.
Note that a scree plot for the BIC can be plotted by calling \texttt{plot(fit)}.
Following the elbow criterion, a three-class solution would also be defensible.
The function \texttt{ic\_weights(tab\_compare)} allows us to compute IC weights; it indicates that, conditional on the set of models, the 4-class model has a posterior model probability of nearly 100\%.

\hypertarget{optional-alternative-model-specifications}{%
\subsection{Optional: Alternative Model Specifications}\label{optional-alternative-model-specifications}}

In the case of confirmatory LCA, the theory would be refuted by strong evidence against the hypothesized model and number of classes.
In the preceding, we only compared the theoretical model against models with different number of classes.
Imagine, however, that a Reviewer argues that variance ought to be freely estimated across classes.
We could compare our theoretical model against their competing model as follows.
Note that we can put two models into a list to compare them.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res\_alt }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(df, }\AttributeTok{classes =} \DecValTok{4}\NormalTok{, }\AttributeTok{variances =} \StringTok{"varying"}\NormalTok{)}
\NormalTok{compare }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(res[[}\DecValTok{4}\NormalTok{]], res\_alt)}
\FunctionTok{table\_fit}\NormalTok{(compare)}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{tab:tabfitcomp}Comparing competing theoretical models}

\begin{tabular}{lllllllll}
\toprule
Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{Parameters} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Entropy} & \multicolumn{1}{c}{prob\_min} & \multicolumn{1}{c}{prob\_max} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{n\_max}\\
\midrule
1 & -1,916.25 & 23 & 3,976.03 & 0.75 & 0.81 & 0.92 & 0.16 & 0.34\\
2 & -1,909.23 & 35 & 4,036.87 & 0.78 & 0.84 & 0.92 & 0.16 & 0.32\\
\bottomrule
\end{tabular}

@@

The alternative model incurs 12 additional parameters for the free variances.
Yet, it has a higher BIC, which indicates that this additional complexity does not outweigh the increase in fit.

To interpret the final class solution, we first reorder the 4-class model by class size.
This helps prevent label switching.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res\_final }\OtherTok{\textless{}{-}} \FunctionTok{mx\_switch\_labels}\NormalTok{(res[[}\DecValTok{4}\NormalTok{]])}
\end{Highlighting}
\end{Shaded}

The 4-class model yielded classes of reasonable size; the largest class comprised 33\%, and the smallest comprised 16\% of cases.
However, the entropy was low, \(S = .75\), indicating poor class separability.
Furthermore, the posterior classification probability ranged from \([.81, .92]\), which means that at least some classes had a high classification error.
We produce a table of the results below.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{table\_results}\NormalTok{(res\_final, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"confint"}\NormalTok{, }\StringTok{"class"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\caption{\label{tab:unnamed-chunk-13}Four-class model results}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Appendix B: Exploratory LPA for Ocean Microplastics},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Appendix B: Exploratory LPA for Ocean Microplastics}
\author{\phantom{0}}
\date{}

@@

This is an example of exploratory latent class analysis (LCA) with continuous indicators, otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using \texttt{tidySEM}.
The present example uses data collected by Alkema as part of a study on ocean microplastics.
The purpose of this study was to provide a more nuanced model for the distribution of different sizes of ocean microplastics than the commonly used normal distribution.
To this end, a mixture of normals was used.
Since there is no theoretical reason to expect a certain number of classes, this is an exploratory LCA.
To view its documentation, run the command \texttt{?tidySEM::alkema\_microplastics} in the R console.
The original analyses are available at \url{https://github.com/cjvanlissa/lise_microplastics}; in this vignette, we take a different approach to the analysis to showcase other possibilities.

@@

To load the data, simply attach the \texttt{tidySEM} package.
For convenience, we assign the variables used for analysis to an object called \texttt{df}.
As explained in the paper, the classes are quite different for lines, films, and fragments.
For this reason, we here only use data from fragments.
The indicators are fragments' length and width.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Load required packages}
\FunctionTok{library}\NormalTok{(tidySEM) }
\FunctionTok{library}\NormalTok{(ggplot2)}
\CommentTok{\# Load data}
\NormalTok{df\_analyze }\OtherTok{\textless{}{-}}\NormalTok{ alkema\_microplastics[alkema\_microplastics}\SpecialCharTok{$}\NormalTok{category }\SpecialCharTok{==} \StringTok{"Fragment"}\NormalTok{, ]}
\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ df\_analyze[ ,}\FunctionTok{c}\NormalTok{(}\StringTok{"length"}\NormalTok{, }\StringTok{"width"}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

@@

As per the best practices, the first step in LCA is examining the observed data.
We use \texttt{tidySEM::descriptives()} to describe the data numerically.
Because all items are categorical, we remove columns for continuous data to de-clutter the table:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ tidySEM}\SpecialCharTok{::}\FunctionTok{descriptives}\NormalTok{(df)}
\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"type"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{, }
\StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{, }\StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
\NormalTok{papaja}\SpecialCharTok{::}\FunctionTok{apa\_table}\NormalTok{(desc, }\AttributeTok{caption =} \StringTok{"Descriptive statistics"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{tabular}{llllllllllll}
\toprule
name & \multicolumn{1}{c}{type} & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{missing} & \multicolumn{1}{c}{unique} & \multicolumn{1}{c}{mean} & \multicolumn{1}{c}{median} & \multicolumn{1}{c}{sd} & \multicolumn{1}{c}{min} & \multicolumn{1}{c}{max} & \multicolumn{1}{c}{skew\_2se} & \multicolumn{1}{c}{kurt\_2se}\\
\midrule
length & numeric & 5605 & 0.00 & 2086 & 2.94 & 2.37 & 1.89 & 1.00 & 69.16 & 137.38 & 2,116.43\\
width & numeric & 5605 & 0.00 & 2079 & 2.00 & 1.62 & 1.11 & 0.20 & 6.76 & 22.23 & 37.08\\
\bottomrule
\end{tabular}

@@

Additionally, we can plot the data.
The \texttt{ggplot2} function \texttt{geom\_density()} is useful to visualize continuous data:

@@

\includegraphics{appendices/plot_gmm_desc.pdf}

The data are correctly coded as \texttt{numeric}.
There are no missing values; if any variables had missing values, we would report an MCAR test with \texttt{mice::mcar()}, and explain that missing data are accounted for using FIML.
Both the table above and the density plot indicate that the data are extremely right-skewed and kurtotic.
With this in mind, it can be useful to transform and rescale the data.
We will use a log transformation.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df\_plot}\SpecialCharTok{$}\NormalTok{Value }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(df\_plot}\SpecialCharTok{$}\NormalTok{Value)}
\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value)) }\SpecialCharTok{+}
  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{Variable)}\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\includegraphics{appendices/plot_gmm_desc_log.pdf}

The log transformation addresses the aforementioned concerns regarding skew and kurtosis.
To confirm this, reshape the data to wide format and examine a scatterplot:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(df\_plot, }\AttributeTok{direction =} \StringTok{"wide"}\NormalTok{, }\AttributeTok{v.names =} \StringTok{"Value"}\NormalTok{)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
\FunctionTok{names}\NormalTok{(df) }\OtherTok{\textless{}{-}} \FunctionTok{gsub}\NormalTok{(}\StringTok{"Value."}\NormalTok{, }\StringTok{""}\NormalTok{, }\FunctionTok{names}\NormalTok{(df), }\AttributeTok{fixed =} \ConstantTok{TRUE}\NormalTok{)}
\FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ length, }\AttributeTok{y =}\NormalTok{ width)) }\SpecialCharTok{+}
  \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{alpha =}\NormalTok{ .}\DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\includegraphics{appendices/plot_gmm_scatter.pdf}

@@

As all variables are continuous, we can use the convenience function \texttt{tidySEM::mx\_profiles()}, which is a wrapper for the generic function \texttt{mx\_mixture()} optimized for continuous indicators.
Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
Its arguments are \texttt{data} and number of \texttt{classes}.
All variables in \texttt{data} are included in the analysis, which is why we first selected the indicator variables.

As this is an exploratory LCA, we will conduct a rather extensive search across model specifications and number of classes.
We will set the maximum number of classes \(K\) to four; depending on the results, we can always choose to increase it later.
We set a seed to ensure replicable results.

As the analysis takes a long time to compute, it is prudent to save the results to disk immediately, so as not to lose them.
For this, we use the function \texttt{saveRDS()}.
We can later use \texttt{res\ \textless{}-\ readRDS("res\_gmm.RData")} to load the analysis from the file.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df,}
                   \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{,}
                   \AttributeTok{variances =} \FunctionTok{c}\NormalTok{(}\StringTok{"equal"}\NormalTok{, }\StringTok{"varying"}\NormalTok{),}
                   \AttributeTok{covariances =} \FunctionTok{c}\NormalTok{(}\StringTok{"zero"}\NormalTok{, }\StringTok{"equal"}\NormalTok{,}
                                   \StringTok{"varying"}\NormalTok{),}
                   \AttributeTok{expand\_grid =} \ConstantTok{TRUE}\NormalTok{)}
\FunctionTok{saveRDS}\NormalTok{(res, }\StringTok{"res\_gmm.RData"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

To compare the fit of the estimated models, we create a model fit table using \texttt{table\_fit()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{table\_fit}\NormalTok{(res)}
\end{Highlighting}
\end{Shaded}

First, we determine whether any models can be disqualified.
There were no indications of convergence problems during estimation, so this is not a reason to disqualify solutions.
Next, we check for local identifiability.
The sample size is \texttt{5605}.
We can calculate the ratio of observations to parameters and append it to the fit table:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit}\SpecialCharTok{$}\NormalTok{par\_ratio }\OtherTok{\textless{}{-}}\NormalTok{ (}\DecValTok{5605}\SpecialCharTok{*}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{n\_min) }\SpecialCharTok{/}\NormalTok{ (fit}\SpecialCharTok{$}\NormalTok{Parameters}\SpecialCharTok{/}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{Classes)}
\NormalTok{fit[ , }\FunctionTok{c}\NormalTok{(}\StringTok{"Name"}\NormalTok{, }\StringTok{"LL"}\NormalTok{, }\StringTok{"Parameters"}\NormalTok{, }\StringTok{"par\_ratio"}\NormalTok{,}
         \StringTok{"BIC"}\NormalTok{, }\StringTok{"Entropy"}\NormalTok{,}
         \StringTok{"prob\_min"}\NormalTok{, }\StringTok{"prob\_max"}\NormalTok{, }
         \StringTok{"n\_min"}\NormalTok{, }\StringTok{"n\_max"}\NormalTok{,}
         \StringTok{"lmr\_p"}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{tab:tabfit}Model fit table. Some columns removed to fit page.}

\begin{tabular}{lllllllll}
\toprule
Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{p} & \multicolumn{1}{c}{par\_ratio} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Ent.} & \multicolumn{1}{c}{p\_min} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{lmr\_p}\\
\midrule
equal var 1 & -8,107.31 & 4 & 1,401.25 & 16,249.14 & 1.00 & 1.00 & 1.00 & NA\\
equal var 2 & -5,211.25 & 7 & 564.00 & 10,482.91 & 0.87 & 0.94 & 0.35 & 0.00\\
equal var 3 & -4,138.27 & 10 & 341.70 & 8,362.85 & 0.83 & 0.88 & 0.20 & 0.00\\
equal var 4 & -3,500.42 & 13 & 246.15 & 7,113.04 & 0.83 & 0.89 & 0.14 & 0.00\\
free var 1 & -8,107.31 & 4 & 1,401.25 & 16,249.14 & 1.00 & 1.00 & 1.00 & NA\\
free var 2 & -5,137.90 & 9 & 501.33 & 10,353.49 & 0.85 & 0.94 & 0.40 & 0.00\\
free var 3 & -4,005.10 & 14 & 374.36 & 8,131.04 & 0.83 & 0.89 & 0.31 & 0.00\\
free var 4 & -3,330.87 & 19 & 245.05 & 6,825.74 & 0.84 & 0.88 & 0.21 & 0.00\\
equal var, equal cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
equal var, equal cov 2 & -3,082.31 & 8 & 432.25 & 6,233.66 & 0.72 & 0.86 & 0.31 & 0.00\\
equal var, equal cov 3 & -3,030.10 & 11 & 256.36 & 6,155.14 & 0.67 & 0.71 & 0.17 & 0.00\\
equal var, equal cov 4 & -3,020.67 & 14 & 220.29 & 6,162.19 & 0.61 & 0.68 & 0.14 & 0.00\\
free var, equal cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
free var, equal cov 2 & -2,544.74 & 10 & 97.00 & 5,175.79 & 0.64 & 0.51 & 0.09 & 0.00\\
free var, equal cov 3 & -2,256.95 & 15 & 71.80 & 4,643.36 & 0.68 & 0.54 & 0.06 & 0.00\\
free var, equal cov 4 & -2,068.70 & 20 & 26.00 & 4,310.02 & 0.63 & 0.52 & 0.02 & 0.00\\
equal var, free cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
equal var, free cov 2 & -2,551.86 & 9 & 107.78 & 5,181.40 & 0.65 & 0.52 & 0.09 & 0.00\\
equal var, free cov 3 & -2,359.39 & 13 & 84.23 & 4,830.99 & 0.68 & 0.56 & 0.07 & 0.00\\
equal var, free cov 4 & -2,173.85 & 17 & 27.53 & 4,494.44 & 0.63 & 0.60 & 0.02 & 0.00\\
free var, free cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
free var, free cov 2 & -2,574.87 & 11 & 406.91 & 5,244.69 & 0.56 & 0.81 & 0.40 & 0.00\\
free var, free cov 3 & -2,111.38 & 17 & 36.00 & 4,369.50 & 0.65 & 0.51 & 0.04 & 0.00\\
free var, free cov 4 & -2,024.10 & 23 & 15.48 & 4,246.73 & 0.62 & 0.50 & 0.02 & 0.00\\
\bottomrule
\end{tabular}

@@

As can be seen from the fit table, the lowest ratio of observations to parameters is 18, which is no cause for concern.

However, note that we have a very large sample, and for many models, the smallest class comprises only a very small percentage of the total sample.
Since the purpose of this analysis is to better represent the distribution of ocean microplastics, we can wonder whether it makes sense to allow for classes that only describe a small percentage of the cases.
We therefore set a lower bound for class size and only consider solutions that capture at least 10\% of the sample.

An interesting characteristic of this data is that the BIC and the entropy are strongly correlated.
The raw correlation between these two metrics is .66, \texttt{cor(fit\$BIC,\ fit\$Entropy)}.
If we omit the 1-class models, for which entropy is technically not defined, the correlation is even as high as .85, \texttt{cor(fit\$BIC{[}!fit\$Classes\ ==\ 1{]},\ fit\$Entropy{[}!fit\$Classes\ ==\ 1{]})}.

This strong correlation indicates that an increase in fit comes with a decrease in class separability.
This illustrates why entropy should not be treated as a model fit criterion.
It also illustrates that criteria for class enumeration should be explicit, because we will likely come to a different decision depending on which criteria are used.

As mentioned before, we drop models with \textless{} 10\% of cases in the smallest class:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit }\OtherTok{\textless{}{-}}\NormalTok{ fit[}\SpecialCharTok{!}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{n\_min }\SpecialCharTok{\textless{}}\NormalTok{ .}\DecValTok{1}\NormalTok{, ]}
\end{Highlighting}
\end{Shaded}

If our strategy is to optimize fit, we can examine the fit table above, or plot a scree plot for the BIC by calling \texttt{plot(fit)}.
Note that, due to the large sample size, all ICs give identical conclusions.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(fit) }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \DecValTok{90}\NormalTok{, }\AttributeTok{vjust =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{hjust=}\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=1\linewidth]{appendices/gmm_plotfit} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-16}
\end{figure}

Looking at the blocks of 1-4 class models for each model specification, it appears that the BIC keeps decreasing with the addition of more classes.
Across the blocks, the BIC keeps decreasing with increasingly complex model specifications.
Similarly, all LMR tests are significant, indicating a preference for more complex models.

Out of the 16 models that remain after removing those with \textless{} 10\% of cases in the smallest class, one model stands out: The 2-class model with free (co)variances.
We thus select this as our final model.

and extract the classification probabilities and model results.
We here request the estimates (\texttt{est}) and standardized estimates \texttt{std\_est}, because the latter allows us to interpret the correlations between length and width.
Note that there is little sense in requesting the standard errors and p-values: With a sample size of 5606, every parameter is significantly different from zero.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res\_bic }\OtherTok{\textless{}{-}}\NormalTok{ res[[}\StringTok{"free var, free cov 2"}\NormalTok{]]}
\NormalTok{cp }\OtherTok{\textless{}{-}} \FunctionTok{class\_prob}\NormalTok{(res\_bic)}
\NormalTok{results }\OtherTok{\textless{}{-}} \FunctionTok{table\_results}\NormalTok{(res\_bic, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"std\_est"}\NormalTok{))}
\NormalTok{results}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{tab:unnamed-chunk-19}Results of a 2-class model with free (co)variances}

\begin{tabular}{lll}
\toprule
label & \multicolumn{1}{c}{est} & \multicolumn{1}{c}{std\_est}\\
\midrule
Means.length & 0.65 & 2.04\\
Means.width & 0.34 & 1.07\\
Variances.length & 0.10 & 1.00\\
Covariances.length.WITH.width & 0.09 & 0.92\\
Variances.width & 0.10 & 1.00\\
Means.length & 1.33 & 3.03\\
Means.width & 0.86 & 1.64\\
Variances.length & 0.19 & 1.00\\
Covariances.length.WITH.width & 0.20 & 0.85\\
Variances.width & 0.28 & 1.00\\
mix2.weights[1,2] & 0.75 & NA\\
\bottomrule
\end{tabular}

@@

Interpreting the results is facilitated by examining a plot of the model and data.
Relevant plot functions are \texttt{plot\_bivariate()}, \texttt{plot\_density()}, and \texttt{plot\_profiles()}.
However, we omit the density plots, because \texttt{plot\_bivariate()} also includes them.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot\_bivariate}\NormalTok{(res\_bic)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=1\linewidth]{appendices/gmm_bivariate_bic} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-22}
\end{figure}

On the diagonal of the bivariate plot are weighted density plots: normal approximations of the density function of observed data, weighed by class probability.
On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.

The bivariate and marginal plots show that the classes are not clearly separable, as also evident from the low entropy.
At the same time however, it is clear that the distributions are non-normal, and the second class accounts for some of this non-normality (there is a smaller `bump' to the right of the mode, which could be the mean of a second normal distribution).
The first class (57\%) accounts for smaller fragments, and the second class (43\%) accounts for some of the right-skew in fragments' length and width.
We can simply label class 1 as \emph{small fragments}, and class 2 as \emph{larger fragments}.

It also appears that the correlation between length and width is stronger for small fragments than for large fragments.
To test the difference, use \texttt{wald\_test(res\_bic,\ hypothesis\ =\ "c11\ =\ c21")}.
Results indicate that the correlation is indeed significantly larger for small fragments (\(r = .92\)) than for larger fragments (\(r = .85\)), \(\chi^2(1) = 11.56, p < .001\).
Thus, small fragments are more coextensive than large fragments.
